---
title: "Hedge Strategy"
author: "Markus Fiordaliso"
date: "10/13/2020"
output: pdf_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(warning = F)
knitr::opts_chunk$set(message = F)
```


# Data

```{r include=F}
# package
library(tidyverse)
library(scales)
library(VIM)
library(lubridate)
library(gridExtra)
```


```{r include=F}
odds_df <- read.csv("/Users/markusfiordaliso/Documents/R/sports_bet/hedge/hedge_data/nfl_20100701_20201012_ml.csv") %>% 
  select(-odds_URL)

```

## Missing Data


Bovada seems to be missing the most data. I will leverage Pinnacle due the legitimacy of the book, coupled with the relatively low levels of missing data.

Some scores are missing. After doing some research on a few of these instances, it looks like sportsbookreview is missing these scores even though the game took place. This can be expected to some degree since I didn't pay for the data. I will factor these out of the analysis
```{r}


missing <- map_df(odds_df, ~sum(is.na(.))) %>% 
  select_if(~any(. > 0))

aggr(odds_df[ , colnames(odds_df) %in% colnames(missing)],
     sortVars = T,
     cex.axis = 0.4,
     oma = c(4,2,1,1))
                           
# remove missing scores from df

odds_df <- odds_df %>% 
  filter(!is.na(away_4Q))
```


## Tidy

```{r}
source("/Users/markusfiordaliso/Documents/R/sports_bet/odds_functions.R")

odds_df <- odds_manipulate_fn(odds_df)

# using pinnacle as the book
sports_book <- "pinnacle"

# filter out missing odds
odds_df <- odds_df %>% 
  filter(!is.na(pinnacle1), !is.na(pinnacle2))

odds_df <- odds_book_remove_fn(odds_df, sports_book)
```


# Odds Overview

Exploring the frequency each odd occurs, as this will give me a better feel on how often I can execute strategies that might target specific odd thresholds. -200:200 is the most frequent odds, which is what I would expect.
```{r fig.width=7}

odds_df <- fav_und_odd_score_fn(odds_df, sports_book)

fav_odd_dist <- odds_df %>% 
  ggplot(aes(x = fav_odd)) +
  geom_histogram(binwidth = 100,
                 aes(y = stat(count) / sum(count)),
                 color = "black",
                 fill = "purple",
                 closed = "right",
                 boundary = 0) +
  scale_x_continuous(breaks = seq(-1000, 0, by = 100)) +
  coord_cartesian(xlim = c(-1000, 0),
                  ylim = c(0, 0.6)) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(y = "Frequency",
       title = "NFL Favorite Odds Frequency",
       caption = "Excluded odds exceeding -1000")

und_odd_dist <- odds_df %>% 
  ggplot(aes(x = und_odd)) +
  geom_histogram(binwidth = 100,
                 aes(y = stat(count) / sum(count)),
                 color = "black",
                 fill = "light blue",
                 closed = "left",
                 boundary = 0) +
  scale_x_continuous(breaks = seq(-100, 1000, by = 100)) +
  coord_cartesian(xlim = c(-200, 1000),
                  ylim = c(0, 0.6)) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(y = "Frequency",
       title = "NFL Underdog Odds Frequency",
       caption = "Excluded odds exceeding +1000")

grid.arrange(fav_odd_dist, und_odd_dist, nrow = 1)
```
Books do not offer true odds since they are the ones who bear the risk. The books take a 5-10% cut on most bets as seen below. However, as the odds gap between teams increases, so does the cut that the book takes. This is to make up for the lack of liquidity on both sides of the bet, and reduce the overall risk that the book takes on. 

```{r fig.width=7}

odds_df <- odds_df %>% 
  mutate(book_spread = abs(ifelse(und_odd > 0, fav_odd + und_odd, fav_odd + 100 + und_odd + 100)) / 100 / 2)

book_spead_view <- odds_df %>% 
  ggplot(aes(x = book_spread)) +
  geom_histogram(aes(y = stat(count) / sum(count)),
                 binwidth = 0.05,
                 boundary = 0,
                 closed = "left",
                 color = "black",
                 fill = "purple") +
  coord_cartesian(xlim = c(0, 1),
                  ylim = c(0, 0.5)) +
  scale_x_continuous(breaks = seq(0, 1, 0.05)) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(y = "Frequency",
       title = "Sports Book Spread per $1")
  

spread_odd_view <- odds_df %>% 
  ggplot(aes(y = und_odd, x = book_spread)) +
  geom_point() +
  geom_smooth() +
  scale_x_continuous(breaks = seq(1, 10, 1)) +
  labs(title = "Book Spread vs Odds Relationship")


grid.arrange(book_spead_view, spread_odd_view, ncol = 2)
```

# Bet the Underdog and Hedge

Picking outright winners can be a challenging feat to generate sustainable profits. Instead I want to focus around hedging a bet during the game to lock in a profit, regardless of outcome. The hypothesis is that if a money line position is taken before a game, at some point the odds will improve and an opportunity will present itself to lock in a profit. Specifically, I want to focus on underdogs as this will provide more upside to guarantee a profit as seen below. However, I will look at both sides when digging into lead changes.


```{r}

odds <- round(c(seq(-300, -100, by = 20), seq(120, 300, by = 20)), 0)
odds_fctr <- ifelse(odds < 0, -100 / odds, odds / 100)

odd_tradeoff <- data.frame(odds,
                           odds_fctr)

odd_tradeoff %>% 
  ggplot(aes(x = factor(odds), y = odds_fctr)) +
  geom_col(color = "black", fill = "purple") +
  scale_y_continuous(labels = percent, breaks = seq(0, 3, 0.5)) +
  labs(x = "Odds",
       y = "Profit",
       title = "Odd & Profit Tradeoff")

```

## Lead Changes

Since I don't have access to historical live odds, I will use lead changes as a proxy to determine changes in odds. The magnitude of the change will be impossible to confirm with this data/approach, but directionally this should give me an idea on how successful this strategy may be. Also, it is important to note that I only have access to scores at the end of each quarter. As a result the true success of this strategy will be limited as I will not be able to capture all lead changes.

Lead changes at any point in the game

```{r}
# create var to see if a lead change happened
odds_df <- odds_df %>% 
  ungroup() %>% 
  mutate(fav_quarters_cum_lead = (fav_1Q_lead + fav_2Q_lead + fav_3Q_lead + fav_4Q_lead),
         und_quarters_cum_lead = (und_1Q_lead + und_2Q_lead + und_3Q_lead + und_4Q_lead))

# summary tables
fav_lead_chg_summary <- odds_df %>% 
  group_by(fav_quarters_cum_lead) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  rename("quarters" = fav_quarters_cum_lead) %>% 
  mutate(freq = count / sum(count),
         type = "Favorite")

und_lead_chg_summary <- odds_df %>% 
  group_by(und_quarters_cum_lead) %>% 
  summarize(count = n()) %>% 
  ungroup() %>% 
  rename("quarters" = und_quarters_cum_lead) %>% 
  mutate(freq = count / sum(count),
         type = "Underdog")

# graph
  

bind_rows(fav_lead_chg_summary, und_lead_chg_summary) %>% 
  ggplot(aes(x = quarters, y = freq, fill = type)) +
  geom_col(position = "dodge",
           color = "black") +
  geom_text(aes(label = percent(freq)),
            position = position_dodge(width = 1),
            vjust = -0.3,
            size = 3) +
  scale_y_continuous(labels = percent) + 
  labs(x = "Quarters",
       y = "Frequency",
       title = "Quarters led per game",
       fill = "")
                 
# continue to validate data before proceeding to ensure lead changes are good
odds_df %>% 
  filter(away_4Q_cum > home_4Q_cum, away_fav == 0, fav_4Q_lead == 1)
```
```{r}
# create range -200:200 to better see if % improve from view above
```

